---
title: "Class 8"
author: "Kevin Tang"
date: "10/25/18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means clustering

Our first example with **kmeans()** function.

```{r}
# Examle plot to see how Rmarkdown works 
plot(1:10,typ="l")
```

Back to kmeans...
# Inspect/print the results
```{r}
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
k <- kmeans(x,centers=2,nstart=20)
k
```
# Q. How many points are in each cluster?
```{r}
k$size
```
# Q. What ‘component’ of your result object details
# - cluster size?
# - cluster assignment/membership?
```{r}
table(k$cluster)
```
# - cluster center?
```{r}
k$centers
```
# Plot x colored by the kmeans cluster assignment and
# add cluster centers as blue points
```{r}
palette(c("blue","red"))
plot(x,col=k$cluster)
points(k$centers,col="green",pch=20,cex=3)
```
# Q. Repeat for k=3, which has the lower tot.withinss? 
```{r}
k3 <- kmeans(x,centers=3,nstart=20)
k3$tot.withinss
k$tot.withinss
```

# Hierarchical clustering in R 

Let's try out the **hclust()** function for Hierarchical clustering in R. This function
needs a distance matrix as input! 

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```
```{r}
plot(hc)
# Draw a line on the dendrogram
abline(h=8,col="red")
# Cut the tre to yield cluster membership vector
cutree(hc,h=8)
cutree(hc,k=2)
```

### A more real example of data clustering
```{r}
# Step 1. Generate some example data for clustering
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
# Step 2. Plot the data without clustering
plot(x)
# Step 3. Generate colors for known clusters
# (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
palette(c("red", "blue", "green"))
plot(x, col=col, pch=20)
```
# Q. Use the dist(), hclust(), plot() and cutree()
# functions to return 2 and 3 clusters
```{r}
d1 <- dist(x)
hc1 <- hclust(d1)
plot(hc1)
g2 <- cutree(hc1, k=2)
g3 <- cutree(hc1, k=3)
plot(x,col=g2,pch=20)
plot(x,col=g3,pch=20)
```
# Q. How does this compare to your known 'col' groups?
```{r}
table(g3, col)
```

```{r}
mydata <- matrix(nrow=100, ncol=10) 
rownames(mydata) <- paste("gene", 1:100, sep="") 
colnames(mydata) <- c( paste("wt", 1:5, sep=""),
 paste("ko", 1:5, sep="") ) 
for(i in 1:nrow(mydata)) {
 wt.values <- rpois(5, lambda=sample(x=10:1000, size=1))
 ko.values <- rpois(5, lambda=sample(x=10:1000, size=1))

 mydata[i,] <- c(wt.values, ko.values)
}
head(mydata)
```
Note the prcomp() functions wants us to take 
```{r}
head(t(mydata))
```
Now lets try to find "structure" in this data with **prcomp()**
```{r}
pca <- prcomp(t(mydata), scale=TRUE)
pca
```
Make a PC plot of PC1 vs PC2. For this I use the '$x' component of our output (i.e. pca$x).
```{r}
plot(pca$x[,1], pca$x[,2]) 
```
Let's see how well our PCs are doing (i.e. how much varaiance from the original data are they capturing)
```{r}
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

pca.var.per
```
Make a "scree-plot" to see the variance captured in each PC
```{r}
barplot(pca.var.per, main="Scree Plot",
 xlab="Principal Component", ylab="Percent Variation")
```
Let's make our plot more useful
```{r}
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)"))
```



